Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies

### 2022-09-07 paper/paper.qmd
rebuild with edits

---
title: Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies
author:
  - name: Dean Marchiori
    email: deanmarchiori@gmail.com
    attributes:
        corresponding: true
abstract: |
  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data in reported cases. This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some as high as 12.5 years. This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes. Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families.
keywords: 
  - NEDMABA
  - SMPD4
#journal: "An awesome journal"
date: last-modified
execute: 
  echo: false 
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature column. Only those features that appeared in more than one case were included to minimise the influence of non-related features in this analysis.

@@

The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous. This was computed in R using the `clValid` package [@clvalid].

@@

The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book]. Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature. A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.

@@

Repeating MCA for the remaining subjects shows the first dimension ($\lambda_! = 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance (@tbl-eig).

@@

Three other smaller satellite clusters exists where highly distinctive features were associated with an individual or family. A full articulation of feature to cluster assignment is given in @tbl-profile.

@@

```{r, warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these comparable metrics."

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (0.69, 0.98). Survival probability after one year at 0.403 (0.252, 0.644) and survival probability at close to ten years of 0.24 (0.11, 0.52).

@@

The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data. This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly identifies the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay. Of significant note is the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. We identify a prolific feature of 'respiratory distress' as an expression more associated with other features such as vocal cord palsy, tracheostomy, tube feeding and GERD which are present in the case reported data, but not well described in HPO. This represents a related but distinct grouping of severe features that associated with this condition.

Other cluster groupings identify outlying features that are expressed in individuals or specific families. Many of the outlying features are specific facial or anatomical dysmorphisms. While course facial features are a noted and core feature, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail. On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation appearing as an isolated feature without the typical microcephaly and hypomyelination. However, there was significant heterogeneity in how feet contractures were reported in clinical case reports. In the data pre-processing, a separate feature was created for reported cases with rocker bottom feet or congenital vertical talus versus those with clubfeet/talipes. In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literature where possible. This is particularly clear as rocker bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature. Another feature that clusters within the core identified features is bilateral cleft lip. This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features. Not all features in the data were included in this analysis, particular if isolated to just one case. In this circumstance, these terms in @tbl-hpo have been labelled as 'Isolated'.

@@

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into the childhood and in some cases into the second decade of life. This work identifies a one year survival rate of 0.403 (0.252, 0.644). A naive estimate, using just the number of deceased patients at this time as proportion of the study population is 0.57. This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability. The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work. Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented other meaningful analysis such as log-rank tests or cox proportional hazard modelling which are likely under-powered when comparing multiple groups.

>>>>

---
title: Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies
author:
  - name: Dean Marchiori
    email: deanmarchiori@gmail.com
    attributes:
        corresponding: true
abstract: |
  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data on reported cases. This research aims to collate and standardise available case reports, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis correctly identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure. Several additional outlying and isolated clinical features were also identified. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years. This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes. This analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition, despite challenges with sparse and inconsistent data. This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
keywords: 
  - NEDMABA
  - SMPD4
#journal: "An awesome journal"
date: last-modified
execute: 
  echo: false 
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature. Only those features that appeared in more than one case were included to minimise the influence of isolated features in this analysis.

@@

The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous. This was computed in `R` using the `clValid` package [@clvalid].

@@

The baseline estimator of all individuals was calculated using the *survival* package in `R` [@survival-book]. Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature. A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* `R` package.

@@

Repeating MCA for the remaining subjects shows the first dimension ($\lambda_1 = 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance (@tbl-eig).

@@

Three other smaller satellite clusters exists where highly distinctive features were associated with a single individual or family. A full articulation of feature to cluster assignment is given in @tbl-profile.

@@

```{r, warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 5 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these comparable metrics."

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98). Survival probability after one year at 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).

@@

The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data. This highlights common sources of variation in the clinical phenotype of the condition itself, rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly identifies a core cluster of the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay. Of significant note is the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. We identify a prolific feature of 'respiratory distress' as an expression more associated with other features such as vocal cord palsy, tracheostomy, tube feeding and GERD which are present in the case reported data, but not well described in HPO. This represents a related but distinct grouping of severe features associated with this condition.

Other smaller clusters identify outlying features that are expressed in individuals or specific families. Many of the outlying features are highly specific facial or anatomical dysmorphisms. While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail. On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation, appearing as an isolated feature without the typical microcephaly and hypomyelination. However, there was significant heterogeneity in how feet contractures were reported in clinical case reports. In the data pre-processing, a separate feature was created for reported cases with rocker bottom feet or congenital vertical talus versus those with clubfeet/talipes. In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of these features, care needs to be taken in distinguishing between similar manifesting features in the literature where possible. This is particularly clear as rocker bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature. Another feature that clusters within the core identified features is bilateral cleft lip. This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features. Not all features in the data were included in this analysis, particular if isolated to just one case. In this circumstance, these terms in @tbl-hpo have been labelled as 'Isolated'.

@@

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into childhood and in some cases into the second decade of life. This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644). A naive estimate, using just the number of deceased patients at this time as proportion of the study population is 0.57. This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability. The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work. Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented other meaningful analysis such as log-rank tests or cox proportional hazard modelling which are likely under-powered when comparing multiple groups.

### 2022-09-07 paper/paper.tex
rebuild with edits

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding the associations of the heterogenous
features present in the clinical phenotype and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data in reported cases. This research aims to
collate and standardise available case reports to analyse and identify
meaningful clusters in the clinical phenotype, and to quantify the
survival probability for children with NEDMABA. To overcome the
challenge of sparse, multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To quantify more accurate survival probabilities, Kaplan-Meier
estimation is formulated to account for censoring in the data. The
analysis identified the classic phenotype for this condition, as well as
one other distinct feature cluster relating to findings of vocal cord
paralysis and respiratory failure. The survival probability for those
affected was found to decline sharply in early infancy with median
survival of 150 days, but with some as high as 12.5 years. This wide
range of outcomes is provisionally associated with variant type however
interpretation of these results are guarded based on very low sample
sizes. Despite challenges with sparse and inconsistent data, this
analysis represents the first of its kind to help describe associations
and trajectories of individuals with this condition and can provide
clinicians and genetic counsellors with better information to aide in
decision making and support for families.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

The MCA coordinates for the first 4 principal dimensions were subject to
clustering in order to find groups of features that were internally
homogeneous and externally heterogeneous. This was computed in R using
the \texttt{clValid} package \citep{clvalid}.

@@

The baseline estimator of all individuals was calculated using the
\emph{survival} package in R \citep{survival-book}. Next the
Kaplan-Meier estimator was stratified by the type of genetic variation
reported in the literature. A log-rank test was performed to detect
differences in the survival curves using methods from
\citet{harrington1982class}, again implemented in the \emph{survival} R
package.

@@

Repeating MCA for the remaining subjects shows the first dimension
(\(\lambda_! = 0.18\)) accounts for 17.88\% of the variance in the data
with the first 4 dimensions of the MCA analysis accounting for 54\% of
the variance (Table~\ref{tbl-eig}).

@@

Three other smaller satellite clusters exists where highly distinctive
features were associated with an individual or family. A full
articulation of feature to cluster assignment is given in
Table~\ref{tbl-profile}.

@@

\caption{\label{fig-clvalid}cluster validation analysis showing various
metrics for cluster evaluation from 2 to 10 clusters. The most optimal
number of clusters are selected by picking the lowest value from one or
many of these comparable metrics.}

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562
days) with median survival time calculated as 150 days. At time 0 the
survival probability was 0.82 (0.69, 0.98). Survival probability after
one year at 0.403 (0.252, 0.644) and survival probability at close to
ten years of 0.24 (0.11, 0.52).

@@

The cluster analysis does not aim to group individuals that are alike,
but rather the clinical features themselves and how these contrasting
clusters account for variation in the data. This highlights common
sources of variation on the clinical phenotype of the condition itself,
rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly
identifies the hallmark features of microcephaly, hypomyelination,
arthrogryposis, simplified gyral pattern and developmental delay. Of
significant note is the presence of an adjacent cluster of related
features that are not as well described in The Human Phenotype Ontology
\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify a
prolific feature of `respiratory distress' as an expression more
associated with other features such as vocal cord palsy, tracheostomy,
tube feeding and GERD which are present in the case reported data, but
not well described in HPO. This represents a related but distinct
grouping of severe features that associated with this condition.

Other cluster groupings identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
specific facial or anatomical dysmorphisms. While course facial features
are a noted and core feature, those with a milder course had clinical
reports that articulated these dysmorphisms in much more detail. On a
similar note, the reporting of `talipes' as a condition which seems
distinct as a source of variation appearing as an isolated feature
without the typical microcephaly and hypomyelination. However, there was
significant heterogeneity in how feet contractures were reported in
clinical case reports. In the data pre-processing, a separate feature
was created for reported cases with rocker bottom feet or congenital
vertical talus versus those with clubfeet/talipes. In practice its not
clear how these diagnoses were arrived at, and given the non-specific
nature of this feature, care needs to be taken in distinguishing between
similar manifesting features in the literature where possible. This is
particularly clear as rocker bottom feet was found closely related in
our analysis to other core features despite the prevalence in HPO being
quite low (2/21), suggesting possible under-reporting or confounded
reporting of this feature. Another feature that clusters within the core
identified features is bilateral cleft lip. This condition has a
reported frequency of just 2/21 (Table~\ref{tbl-hpo}) and may be an
emerging feature for a more typical presentation given its close
relationship with other classic features. Not all features in the data
were included in this analysis, particular if isolated to just one case.
In this circumstance, these terms in Table~\ref{tbl-hpo} have been
labelled as `Isolated'.

@@

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into the
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (0.252, 0.644). A naive
estimate, using just the number of deceased patients at this time as
proportion of the study population is 0.57. This highlights the
importance of applying a survival analysis technique that incorporates
censoring so as not to erroneously overestimate the survival
probability. The current literature suggests the type of gene variant
has an impact on survival time, which seems to be supported in this
work. Through a comparison of multiple survival curves, stratified by
variation type, the limited sample size prevented other meaningful
analysis such as log-rank tests or cox proportional hazard modelling
which are likely under-powered when comparing multiple groups.

>>>>

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This research aims to
collate and standardise available case reports, to analyse and identify
meaningful clusters in the clinical phenotype, and to quantify the
survival probability for children with NEDMABA. To overcome the
challenge of multidimensional data on very few subjects, we employ
Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To quantify more accurate survival probabilities, Kaplan-Meier
estimation is formulated to account for censoring in the data. The
analysis correctly identified the classic phenotype for this condition,
as well as one other distinct feature cluster relating to findings of
vocal cord paralysis and respiratory failure. Several additional
outlying and isolated clinical features were also identified. The
survival probability for those affected was found to decline sharply in
early infancy with median survival of 150 days, but with some surviving
as long as 12.5 years. This wide range of outcomes is provisionally
associated with variant type however interpretation of these results are
guarded based on very low sample sizes. This analysis represents the
first of its kind to help describe associations and trajectories of
individuals with this condition, despite challenges with sparse and
inconsistent data. This analysis can provide clinicians and genetic
counsellors with better information to aide in decision making and
support for families with this rare condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

The MCA coordinates for the first 4 principal dimensions were subject to
clustering in order to find groups of features that were internally
homogeneous and externally heterogeneous. This was computed in
\texttt{R} using the \texttt{clValid} package \citep{clvalid}.

@@

The baseline estimator of all individuals was calculated using the
\emph{survival} package in \texttt{R} \citep{survival-book}. Next the
Kaplan-Meier estimator was stratified by the type of genetic variation
reported in the literature. A log-rank test was performed to detect
differences in the survival curves using methods from
\citet{harrington1982class}, again implemented in the \emph{survival}
\texttt{R} package.

@@

Repeating MCA for the remaining subjects shows the first dimension
(\(\lambda_1 = 0.18\)) accounts for 17.88\% of the variance in the data
with the first 4 dimensions of the MCA analysis accounting for 54\% of
the variance (Table~\ref{tbl-eig}).

@@

Three other smaller satellite clusters exists where highly distinctive
features were associated with a single individual or family. A full
articulation of feature to cluster assignment is given in
Table~\ref{tbl-profile}.

@@

\caption{\label{fig-clvalid}cluster validation analysis showing various
metrics for cluster evaluation from 2 to 5 clusters. The most optimal
number of clusters are selected by picking the lowest value from one or
many of these comparable metrics.}

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562
days) with median survival time calculated as 150 days. At time 0 the
survival probability was 0.82 (95\% CI: 0.69, 0.98). Survival
probability after one year at 0.403 (95\% CI: 0.252, 0.644) and survival
probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).

@@

The cluster analysis does not aim to group individuals that are alike,
but rather the clinical features themselves and how these contrasting
clusters account for variation in the data. This highlights common
sources of variation in the clinical phenotype of the condition itself,
rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly
identifies a core cluster of the hallmark features of microcephaly,
hypomyelination, arthrogryposis, simplified gyral pattern and
developmental delay. Of significant note is the presence of an adjacent
cluster of related features that are not as well described in The Human
Phenotype Ontology \citep{kohler2021human} as shown in
Table~\ref{tbl-hpo}. We identify a prolific feature of `respiratory
distress' as an expression more associated with other features such as
vocal cord palsy, tracheostomy, tube feeding and GERD which are present
in the case reported data, but not well described in HPO. This
represents a related but distinct grouping of severe features associated
with this condition.

Other smaller clusters identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
highly specific facial or anatomical dysmorphisms. While coarse facial
features are a noted and core presentation, those with a milder course
had clinical reports that articulated these dysmorphisms in much more
detail. On a similar note, the reporting of `talipes' as a condition
which seems distinct as a source of variation, appearing as an isolated
feature without the typical microcephaly and hypomyelination. However,
there was significant heterogeneity in how feet contractures were
reported in clinical case reports. In the data pre-processing, a
separate feature was created for reported cases with rocker bottom feet
or congenital vertical talus versus those with clubfeet/talipes. In
practice its not clear how these diagnoses were arrived at, and given
the non-specific nature of these features, care needs to be taken in
distinguishing between similar manifesting features in the literature
where possible. This is particularly clear as rocker bottom feet was
found closely related in our analysis to other core features despite the
prevalence in HPO being quite low (2/21), suggesting possible
under-reporting or confounded reporting of this feature. Another feature
that clusters within the core identified features is bilateral cleft
lip. This condition has a reported frequency of just 2/21
(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
presentation given its close relationship with other classic features.
Not all features in the data were included in this analysis, particular
if isolated to just one case. In this circumstance, these terms in
Table~\ref{tbl-hpo} have been labelled as `Isolated'.

@@

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). A
naive estimate, using just the number of deceased patients at this time
as proportion of the study population is 0.57. This highlights the
importance of applying a survival analysis technique that incorporates
censoring so as not to erroneously overestimate the survival
probability. The current literature suggests the type of gene variant
has an impact on survival time, which seems to be supported in this
work. Through a comparison of multiple survival curves, stratified by
variation type, the limited sample size prevented other meaningful
analysis such as log-rank tests or cox proportional hazard modelling
which are likely under-powered when comparing multiple groups.
